VariantSpark: Population Scale Clustering of Genotype Information

### 2015-09-11 doc/publications/journal2014/paper.tex
fixed more comments

Clustering the samples is also fastest in \variantSpark{} (1min and 20secs), despite \variantSpark{} and ADAM both using {\sc Spark}'s MLlib \kMeans{} implementation, see Table~\ref{fivewaycomparison}.
% DENIS: how sure are we that this is not just a fluke, e.g. I typically repeat this 5 times and report the average and std/ste ...
We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient.
Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise 8 cores. 
Despite this, both were an order of magnitude slower than \variantSpark{} and ADAM, with python requiring 11mins and 29secs and R requiring 7mins and 25secs.
Our Hadoop implementation performed the worst with 14mins and 23secs.
The substantially increased runtime is due the dataset (chromosome 22) being small enough to fit into memory. This enables Python and R to store the output from each \kMeans{} iteration in memory. Hadoop, however, writes the entire output of each iteration to disk.
Although this feature allows for greater scalability, as memory availability is less of an issue with huge datasets, the bottleneck introduced from disk IO increases processing time notably.
{\sc Spark} overcomes this limitation by performing in-memory caching of datasets. Caching virtually eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where at least a portion of the data can still be cached in memory.

@@

We run this comparison on our in-house Hadoop cluster using 64 cores, although this was limited to 40 cores with 100\% of the genome due to memory availability.
Pre-processing the VCF files takes approximately 12 minutes, 19 minutes, 27 minutes and 40 minutes for 20\%, 40\%, 60\% and 100\% of the genome, respectively.
In each case, the memory required does not exceed a modest 2GB per executor, even for 100\% of the genome. 
Clustering the data takes 70 minutes, 139 minutes, 213 minutes and 884 mins, respectively.
For clustering, the memory requirements increase approximately linearly with 6GB, 12GB, 17GB and 24GB, respectively, per executor. 
This substantial increase in memory required is a result of the increased dimensionality of the data causing the \kMeans{} centroids, which MLlib stores as dense vectors, requiring excessive memory.  
% DENIS: This conflicts with the 30 seconds gain over ADAM which you attribute to our clustering using sparse vectors,
% AIDAN: Only centroids are dense. Reworded to explain this
% TODO: is it farther ?
% AIDAN: I had to check (http://www.quickanddirtytips.com/education/grammar/further-versus-farther) but I believe it is further.
While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.

@@

%One factor which results in these differing memory requirements is the sparsity of the vectors used to store the variants. Due to their sparse nature (the abundance of 0s) we can store them efficiently as Sparse Vectors in our pre-processing stage. However, to cluster these vectors, the \kMeans{} algorithm converts these sparse vectors to dense vectors. This is much less memory-efficient, as dense vectors explicitly define zero values (which are excluded from Sparse Vectors so don't require any memory), considerably increasing the memory required.
%So although we demonstrate the potential of \kMeans{} clustering to scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, or for example, easily deal with more alleles.

@@

% TODO add number
Again we cluster the individuals allowing X clusters and compare the assigned labels to the same number of self-reported population labels.
The observed ARI is 0.1. 
This poor result may be due to the population annotation being too detailed and diverse especially so for self-reported and hence potential uncertain ancestry.
%Some observations are that the population groups to diverse in the PGP data. 
%Each individual has metadata stating which country each of their grandparents are from, and only 40 individuals have all four grandparents from the same country.
%Also, being open to the public, the data may be more prone to error. For some individuals, the stated country of their grandparents changes between their updates. This could indicate some uncertainty about their origins, and hence, correct population.
% TODO provide mapping
We therefore only include individuals where all grandparents were reported to be from the same country and group them into their approximate super-populations (see Supplemental Table X). 
This new dataset comprises XX individuals.
%To try improve on these results, we only include the individuals with their grandparents were from the same country.
%We then grouped them into their approximate super-populations. 
Clustering these individuals results in an ARI of 0.29.

%DENIS: If you see an increase in accuracy above we can hypothesise here that 985,790 variants are not sufficient... otherwise a sentence around the population mix in potential real world clinical cohorts....

>>>>

Clustering the samples is also fastest in \variantSpark{} (1min and 20secs), despite \variantSpark{} and ADAM both using {\sc Spark}'s MLlib \kMeans{} implementation, see Table~\ref{fivewaycomparison}.
% DENIS: how sure are we that this is not just a fluke, e.g. I typically repeat this 5 times and report the average and std/ste ...
% AIDAN: This result was consistent. But it was on the Azure cluster so I can't get std/ste..
We attribute the 30\% speedup over ADAM (1min and 52secs) to \variantSpark{} converting the VCF files to sparse vectors, whereas ADAM creates dense vectors which are less memory efficient.
Multithreading \kMeans{} clustering is natively supported by both R and Python, we are therefore able to utilise 8 cores. 
Despite this, both were an order of magnitude slower than \variantSpark{} and ADAM, with python requiring 11mins and 29secs and R requiring 7mins and 25secs.
Our Hadoop implementation performed the worst with 14mins and 23secs.
The substantially increased runtime is due the dataset (chromosome 22) being small enough to fit into memory. This enables Python and R to store the output from each \kMeans{} iteration in memory. Hadoop, however, writes the entire output of each iteration to disk.
Although this feature allows for greater scalability, as memory availability is less of an issue with huge datasets, the bottleneck introduced from disk IO increases processing time notably.
{\sc Spark} overcomes this limitation by performing in-memory caching of datasets. Caching virtually eliminates the disk IO bottleneck on smaller datasets, and minimises the bottleneck on larger datasets where at least a portion of the data can still be cached in memory.

@@

We run this comparison on our in-house Hadoop cluster using 64 cores, although this was limited to 40 cores with 100\% of the genome due to memory availability.
Pre-processing the VCF files takes approximately 12 minutes, 19 minutes, 27 minutes and 40 minutes for 20\%, 40\%, 60\% and 100\% of the genome, respectively.
In each case, the memory required does not exceed a modest 2GB per executor, even for 100\% of the genome. 
Clustering the data takes 70 minutes, 139 minutes, 213 minutes and 884 mins, respectively.
For clustering, the memory requirements increase approximately linearly with 6GB, 12GB, 17GB and 24GB, respectively, per executor. 
This substantial increase in memory required is a result of the increased dimensionality of the data causing the \kMeans{} centroids, which MLlib stores as dense vectors, requiring excessive memory.  
% DENIS: This conflicts with the 30 seconds gain over ADAM which you attribute to our clustering using sparse vectors,
% AIDAN: Only centroids are dense. Reworded to explain this
% TODO: is it farther ?
% AIDAN: I had to check, but I believe it is further? (http://www.quickanddirtytips.com/education/grammar/further-versus-farther)
While \kMeans{} clustering can scale to 100\% of the genome, machine learning algorithms that can deal with sparse vectors would potentially be able to scale much further, for example process more alleles.

@@

@@

%AIDAN: Changed this section..
Again we cluster the individuals allowing 5 clusters and compare the assigned labels to the same number of self-reported `Race/ethnicity' labels (`White', `Hispanic or Latino', `Asian', `Black or African American' and `American Indian').
The observed ARI is negative, indicating the clustering is completely random. 
This poor result may be due to the labels being too broad and being a blanket label for a diverse range of individuals. For example, the `White' label includes individuals with grandparents from United States to Syrian Arab Republic to Bulgaria.
The dataset is also very one-sided with 433 out of the 478 (over 90\%) of the individuals being classified as white, even though geographical location of their grandparents are very diverse.
%This poor result may be due to the population annotation being too detailed and diverse, especially so for self-reported, and hence potential uncertain ancestry.
% TODO provide mapping
We therefore only include individuals where all grandparents were reported to be from the same country and group them into their approximate super-populations (see Supplemental Table X). 
This new dataset comprises 36 individuals.
Clustering these individuals results in an ARI of 0.29.

%DENIS: If you see an increase in accuracy above we can hypothesise here that 985,790 variants are not sufficient... otherwise a sentence around the population mix in potential real world clinical cohorts.... 
Although we see an increase, the relatively small ARI is likely due to the diversity in this real world clinical cohort across the small sample size. Although our filter removes individuals with a mixed background, we are left with only 36 individuals from 23 different countries.
Applying a super-population mask, based on geographical distance, to these individuals reduces the number of groups from 23 to 5, however this grouping doesn't take migrations that occurred prior to their grandparents into account.
